---
title: "PML-Prediction"
author: "Ian Gonsalves"
date: "1/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# setwd("~/Desktop/Cour-PML")
setwd("~/Downloads/PML")
```

## Introduction  
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).   

## Data
The training data for this project are available here:  
[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)  
The test data is available here:  
[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)  
The data for this project comes from this source: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har). If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.  

  

## Load Libraries

Note: Please **install packages** if they don't already exist on your system
Loading Libraries for this project, which you should install and load them in your working environment.  
```{r warning=FALSE, error=FALSE, echo=FALSE}
# install.packages("DMwR")
library(DMwR)
library(rattle)
library(caret)
library(rpart)
library(rpart.plot)
library(skimr)
library(corrplot)
library(randomForest)
library(RColorBrewer)
```  

## Reproduction of Project Outcome
We set a seed, so anyone reviewing the code can reproduce the results
```{r warning=FALSE, error=FALSE}
set.seed(9999)
```  

## Getting Data  
Setup working directory.  
```{r warning=FALSE, error=FALSE}
# setwd("~/Downloads/GitHub/Cour-PML/Project")
setwd("~/Downloads/PML")
```  

Get dataset to the `data` folder in the current working directory.  
```{r warning=FALSE, error=FALSE}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile = trainFile, method = "curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile = testFile, method = "curl")
}
rm(trainUrl)
rm(testUrl)
```  

## Reading Data  
Load training and testing data into dataframes  
```{r warning=FALSE, error=FALSE}
trainData <- read.csv(trainFile)
testData <- read.csv(testFile)
dim(trainData)
dim(testData)
rm(trainFile)
rm(testFile)
```  
The training data set contains `r dim(trainData)[1]` observations and `r dim(trainData)[2]` variables, while the testing data set contains `r dim(testData)[1]` observations and `r dim(testData)[2]` variables. The `classe` variable in the training set is the outcome to predict.  


## Cleaning Data  
Remove missing values / Remove insignificant variables.  

1. We clean the <b>Near Zero Variance</b> Variables.  
```{r warning=FALSE, error=FALSE, echo=FALSE}
NZV <- nearZeroVar(trainData, saveMetrics = TRUE)
head(NZV, 20)
training01 <- trainData[, !NZV$nzv]
testing01 <- testData[, !NZV$nzv]
dim(training01)
dim(testing01)
rm(trainData)
rm(testData)
rm(NZV)
```  

2. Removing some columns of the dataset that do not contribute much to the accelerometer measurements.  
```{r warning=FALSE, error=FALSE}
regex <- grepl("^X|timestamp|user_name", names(training01))
training <- training01[, !regex]
testing <- testing01[, !regex]
rm(regex)
rm(training01)
rm(testing01)
dim(training)
dim(testing)
```  

3. Removing variables with contain `NA' values.  
```{r warning=FALSE, error=FALSE}
cond <- (colSums(is.na(training)) == 0)
training <- training[, cond]
testing <- testing[, cond]
rm(cond)
```  

Training set contains `r dim(training)[1]` observations and `r dim(training)[2]` variables
Testing set contains `r dim(testing)[1]` observations and `r dim(testing)[2]` variables.  

Plot a matrix of the features correlation in the Training set.  
```{r warning=FALSE, error=FALSE}
corrplot(cor(training[, -length(names(training))]), method = "color", tl.cex = 0.5)
```  

## Break up our data into a Training / Validation Set  
We subset into training data (75%) and a validation data (25%). 
Validation used for cross validation.  
```{r warning=FALSE, error=FALSE}
set.seed(9999)
inTrain <- createDataPartition(training$classe, p = 0.75, list = FALSE)
validation <- training[-inTrain, ]
training <- training[inTrain, ]
rm(inTrain)
```  

The **Dataset** now consists of `r dim(training)[2]` variables with the observations divided as following:  
1. Training Data: `r dim(training)[1]` observations.  
2. Validation Data: `r dim(validation)[1]` observations.  
3. Testing Data: `r dim(testing)[1]` observations.  


## Predictive Modeling Phase  



### 1: Decision Tree  
Build a <b>Decision Tree</b> based Predictive Model for activity recognition.  
```{r warning=FALSE, error=FALSE}
modelDT <- rpart(classe ~ ., data = training, method = "class")
prp(modelDT)
```  

Review performance of newly created Decision Tree prediction model on the <b>Validation</b> data.  
```{r warning=FALSE, error=FALSE, echo=FALSE}
predictDT <- predict(modelDT, validation, type = "class")
confusionMatrix(validation$classe, predictDT)
accuracy <- postResample(predictDT, validation$classe)
sampErr <- 1 - as.numeric(confusionMatrix(validation$classe, predictDT)$overall[1])
rm(predictDT)
rm(modelDT)
```  

The Estimated Accuracy of the Random Forest Model is `r accuracy[1]*100`% and the Estimated Out-of-Sample Error is `r sampErr*100`%.  

### 2: Random Forest
We fit a predictive model for activity recognition using <b>Random Forest</b> algorithm because it automatically selects important variables and is robust to correlated covariates & outliers in general.  
We will use <b>5-fold cross validation</b> when applying the algorithm.  
```{r modelRF, warning=FALSE, error=FALSE}
modelRF <- train(classe ~ ., data = training, method = "rf", trControl = trainControl(method = "cv", 5), ntree = 250)
modelRF
```  

Now, we estimate the performance of the model on the <b>validation</b> data set.  
```{r warning=FALSE, error=FALSE}
predictRF <- predict(modelRF, validation)
confusionMatrix(validation$classe, predictRF)
accuracy <- postResample(predictRF, validation$classe)
sampErr <- 1 - as.numeric(confusionMatrix(validation$classe, predictRF)$overall[1])
rm(predictRF)
```  

The Estimated Accuracy of the Random Forest Model is `r accuracy[1]*100`% and the Estimated Out-of-Sample Error is `r sampErr*100`%.  
Random Forests yielded better Results, as expected!  



## Perform Prediction on the Testing Data  
Now, we apply the <b>Random Forest</b> model to the original testing data set downloaded from the data source. We remove the problem_id column first.  
```{r warning=FALSE, error=FALSE}
ncol(testing)
rm(accuracy)
rm(sampErr)
# predict(modelRF, testing[, -length(names(testing))])
predict(modelRF, testing[, -ncol(testing)])
```  

## Output for Prediction quiz for the 20 questions  
```{r warning=FALSE, error=FALSE}
predictOut = function(x){
  n = length(x)
  for(i in 1:n){
    # filenames = paste0("./PredictionQuiz/Question_",i,".txt")
    filenames = paste0("./question_",i,".txt")
    write.table(x[i], file = filenames, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}
```  
Generating the Files.  
```{r warning=FALSE, error=FALSE}
predictOut(predict(modelRF, testing[, -ncol(testing)]))
rm(modelRF)
rm(training)
rm(testing)
rm(validation)
rm(predictOut)
```  